---
title: "IOT Network Intrusion Detection Analysis"
subtitle: "Q3 : Which combinations of dimensions is responsible for the different type of attack ?"
author: 
  - name: "ViZZards"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project aims to analyze and visualize network attacks within Internet of Things (IoT) environments using the RT-IoT 2022 dataset, which includes network traffic data from both normal and malicious activities across various IoT devices. By examining metrics like protocol usage, bandwidth, payload size, and flow characteristics, the study seeks to identify distinctive patterns associated with different types of attacks. These insights will contribute to developing more effective intrusion detection strategies, enhancing cybersecurity within IoT systems. Ethical considerations are minimal, as the dataset excludes sensitive information."

format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
toc: true
editor: visual
code-annotations: hover
execute:
  warning: false
  error: false
  message: false
---

## Libraries and Setup

```{r}
#| label: load-pkgs
#| message: false
if (!require(pacman)) 
  install.packages("pacman")

pacman::p_load(
  # Core data manipulation and visualization
  tidyverse,
  glue,
  scales,
  ggplot2,
  grid,
  readr,
  corrplot,
  dplyr,
  
  
  # Additional visualization packages
  viridis,
  RColorBrewer,
  
  # Data manipulation
  data.table,
  tibble,
  
  # Additional utilities
  stringr,
  lubridate,
  forcats,
  purrr,
  tidyr,
  magrittr
)


```



## Dataset

```{r}
#| label: load-dataset
#| message: false


data <- read_csv("data.csv")

# Assign distinct colors to each attack type
unique_attacks <- unique(data$Attack_type)
attack_colors <- hue_pal()(length(unique_attacks))

# Filter dataset to exclude NMAP attacks
filtered_data <- data %>%
  filter(!grepl("NMAP", Attack_type, ignore.case = TRUE))

```

## Bandwidth Analysis

```{r}
#| label: Badwidth Analysis
#| message: false

# Remove extreme outliers (top 1%)
bandwidth_data <- filtered_data %>%
  filter(flow_pkts_per_sec < quantile(flow_pkts_per_sec,
                                      0.99,
                                      na.rm = TRUE))

# Create the boxplot with improved scaling and layout
ggplot(bandwidth_data, aes(x = Attack_type, 
                           y = flow_pkts_per_sec, 
                           fill = Attack_type)) +
  geom_boxplot(outlier.shape = NA) + # Suppress individual outliers 
  scale_fill_manual(values = attack_colors) + # Custom colors for attack types
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10), # Adjust x-axis
    axis.title = element_text(size = 12),
    legend.position = "right"
  ) +
  scale_y_log10(labels = scales::comma) + # Apply log10 scale 
  labs(
    title = "Flow Packets per Second by Attack Type (Bandwidth Analysis)",
    x = "Attack Type",
    y = "Flow Packets per Second (Log Scale)",
    fill = "Attack Type"
  )


```
### Analysis :- 

**Variable Interaction:**

High values of flow_pkts_per_sec (e.g., in DOS_SYN_Hping) indicate 
rapid packet flooding, overwhelming network bandwidth.
Other attack types, like DDOS_Slowloris, exhibit low and steady 
flow packet rates, emphasizing a strategic, resource-depleting attack style.

**Inference:**

Attacks like DOS_SYN_Hping utilize high-bandwidth strategies,
while others like ARP_poisoning rely on precise, low-rate operations.
This interaction suggests that bandwidth-intensive attacks overwhelm 
network resources rapidly, while low-rate attacks focus on exploiting 
system vulnerabilities.


## Inter-Arrival Time Analysis

```{r}
#| label: Inter-Arrival Time Analysis 
#| message: false

# Remove extreme outliers (top 1%)
iat_data <- filtered_data %>%
  filter(fwd_iat.avg < quantile(fwd_iat.avg, 0.99, na.rm = TRUE))

# Create the boxplot with improved formatting and log scale
ggplot(iat_data, aes(x = Attack_type, y = fwd_iat.avg, fill = Attack_type)) +
  geom_boxplot(outlier.shape = NA) + # Suppress individual outliers
  scale_fill_manual(values = attack_colors) + # Custom colors for attack types
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10), 
    axis.title = element_text(size = 12),
    legend.position = "right"
  ) +
  scale_y_log10(labels = scales::comma) + # Apply log10 scale for y-axis
  labs(
    title = "Average Forward Inter-Arrival Time by Attack Type",
    x = "Attack Type",
    y = "Forward Inter-Arrival Time (Log Scale)",
    fill = "Attack Type"
  )


```
### Analysis :- 

**Variable Interaction:**

fwd_iat.avg (average inter-arrival time) and its distribution differ
across attack types.

DDOS_Slowloris shows high inter-arrival times,
reflecting deliberate delays.

MQTT_Publish exhibits tightly packed inter-arrival times, 
indicating efficient payload delivery.

**Inference:**

Slowloris attacks exploit resource-holding behaviors 
with prolonged intervals, interacting with server timeout mechanisms.

Low inter-arrival time attacks (e.g., MQTT_Publish) use speed to 
avoid detection and maximize payload delivery, highlighting temporal 
manipulation as a key tactic.





## Payload Behavior Analysis

```{r}
#| label: Payload Behavior Analysis
#| message: false

# Remove extreme outliers (top 1%)
payload_data <- filtered_data %>%
  filter(fwd_pkts_payload.avg < quantile(fwd_pkts_payload.avg, 
                                         0.99,
                                         na.rm = TRUE))

# Create the boxplot with focused y-axis range
ggplot(payload_data, aes(x = Attack_type,
                         y = fwd_pkts_payload.avg,
                         fill = Attack_type)) +
  geom_boxplot(outlier.shape = NA) + # Suppress individual outliers
  scale_fill_manual(values = attack_colors) + # Custom colors for attack types
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, 
                               hjust = 1, 
                               size = 10), 
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  coord_cartesian(ylim = c(10, 100)) + # Focus on y-axis range 10 to 100
  labs(
    title = "Average Forward Packets Payload by Attack Type (Focused Range)",
    x = "Attack Type",
    y = "Forward Packets Payload (Average)",
    fill = "Attack Type"
  )



```
### Analysis:-

**Variable Interaction:**

Variables like fwd_pkts_payload.avg and fwd_pkts_payload.max 
interact to characterize payload distribution.

Wipro_bulb shows a wide payload range, indicating varied message 
sizes, while DDOS_Slowloris clusters at smaller payloads.

**Inference:**

Larger payload ranges indicate dynamic and adaptable attacks 
(e.g., Wipro_bulb), while smaller ranges (e.g., Slowloris) rely 
on consistent patterns to exhaust resources.

Payload variability can be a signature of attack intent 
(e.g., reconnaissance vs. volumetric).



## Heatmaps for Each Attack Type

```{r}
#| label: Heatmaps for Each Attack Type
#| message: false

# Select only bandwidth, packet size, and flow-related variables
heatmap_vars <- c(
  "fwd_pkts_tot", "bwd_pkts_tot", "fwd_data_pkts_tot", "bwd_data_pkts_tot",
  "fwd_pkts_per_sec", "bwd_pkts_per_sec", "flow_pkts_per_sec",
  "fwd_pkts_payload.avg", "fwd_pkts_payload.min", "fwd_pkts_payload.max",
  "flow_duration"
)

# Loop through each unique attack type
for (attack in unique(filtered_data$Attack_type)) {
  # Filter data for the specific attack type
  attack_data <- filtered_data %>% filter(Attack_type == attack)
  
  # Ensure there are enough observations to compute correlations
  if (nrow(attack_data) < 5) next
  
  # Subset only the relevant columns present in attack_data
  relevant_vars <- intersect(heatmap_vars, names(attack_data))
  
  # Remove variables with constant values or too many missing values
  valid_vars <- sapply(attack_data[, relevant_vars, drop = FALSE], function(col) {
    var(col, na.rm = TRUE) > 0 & sum(is.na(col)) / length(col) < 0.5
  })
  
  # Subset the attack_data to valid variables only
  attack_data <- attack_data[, names(valid_vars)[valid_vars], drop = FALSE]
  
  # Calculate correlation matrix, excluding NA values
  if (ncol(attack_data) > 1) {
    corr_matrix <- cor(attack_data, use = "pairwise.complete.obs")
    
    # Replace NA correlations with 0 (optional)
    corr_matrix[is.na(corr_matrix)] <- 0
    
    # Create the heatmap
    heatmap <- ggplot(melt(corr_matrix), aes(Var1, Var2, fill = value)) +
      geom_tile(color = "white") +
      scale_fill_gradient2(
        low = "darkblue", mid = "white", high = "darkred", midpoint = 0, 
        limit = c(-1, 1), name = "Correlation"
      ) +
      labs(
        title = paste("Correlation Heatmap for", attack),
        x = "Variables",
        y = "Variables"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
      )
    
    # Print the heatmap
    print(heatmap)
  }
}


```

### Analysis :-

**1. ARP_Poisoning**

**Variable Interaction:**
Strong correlation between bwd_pkts_tot and flow_pkts_per_sec 
suggests balanced bidirectional traffic.

**Inference:**
ARP poisoning exploits bidirectional communication, emphasizing 
the importance of packet symmetry in its success.


**2. DDOS_Slowloris**

**Variable Interaction:**
Weak correlations across all bandwidth and flow metrics.

**Inference:**
Interaction between low bandwidth, small payloads, and 
long durations indicates a resource-depleting strategy 
without significant data transfer.

**3. DOS_SYN_Hping**

**Variable Interaction:**
High correlation between fwd_pkts_per_sec and fwd_pkts_tot.

**Inference:**
This attack relies on large volumes of forward packets to 
overwhelm servers, showing a direct relationship between 
packet frequency and attack intensity.

**4. Metasploit_Brute_Force_SSH**

**Variable Interaction:**
Correlation between flow_duration and fwd_pkts_payload.avg 
highlights persistence.

**Inference:**
The attack uses sustained sessions with consistent payloads,
emphasizing authentication cracking through repetitive probes.

**5. MQTT_Publish**

**Variable Interaction:**
High correlation between fwd_pkts_payload.max and 
fwd_pkts_per_sec reflects optimized payload delivery.

**Inference:**
MQTT_Publish exploits efficiency in payload delivery, 
highlighting its IoT application focus.

**6. Thing_Speak**

**Variable Interaction:**
Weak overall correlations, but moderate between 
bwd_pkts_tot and fwd_pkts_payload.avg.

**Inference:**
Thing_Speak operates with minimal but consistent 
bidirectional traffic, leveraging IoT protocols for 
covert data exchanges.

**7. Wipro_Bulb**

**Variable Interaction:**
Strong correlation between fwd_pkts_payload.max 
and flow_pkts_per_sec.

**Inference:**
The attack utilizes high variability in payload and 
packet flow, reflecting erratic patterns aimed at 
dynamic IoT environments.


## Synthesis of Variable Interactions Across Attack Types:-

**1. Bandwidth Utilization:**

High flow_pkts_per_sec correlates with volumetric attacks 
like DOS_SYN_Hping, which focus on resource saturation.
Low but steady flow rates signify targeted, resource-depleting 
attacks (e.g., DDOS_Slowloris).

**2. Temporal Manipulation:**

fwd_iat.avg is a defining characteristic for attacks 
like Slowloris, which exploit high inter-arrival times 
to hold server resources.

**3. Payload Dynamics:**

Variability in fwd_pkts_payload.max reflects attack 
adaptability. Static payloads indicate systematic 
attacks (e.g., Metasploit_Brute_Force_SSH), while 
dynamic payloads (e.g., Wipro_bulb) signify flexible, 
IoT-specific threats.

**4. Bidirectional Traffic:**

Correlations between bwd_pkts_tot and fwd_pkts_tot 
highlight the reliance on bidirectional traffic for 
attacks like ARP_poisoning, which exploit network 
routing vulnerabilities.

**5.Attack Fingerprinting:**

Each attack type displays unique interaction patterns
among variables, creating distinct signatures. These
can inform the development of targeted anomaly detection
systems.


## Conclusion

The interaction of variables such as bandwidth, payload, 
and inter-arrival time during an attack provides critical 
insights into the mechanisms and intent of various attack 
types on IoT devices. High bandwidth and rapid packet flows 
signify volumetric attacks, while prolonged inter-arrival 
times suggest resource-exhausting tactics. Variable correlations, 
like those observed in payload size and flow rates, underline 
the adaptability and stealth of IoT-targeted attacks.

**Significance: Understanding these interactions allows for:**

**Attack Fingerprinting:** Identifying unique behavioral signatures 
for each attack.

**Anomaly Detection:** Designing systems that leverage these 
patterns for early detection.

**IoT-Specific Defenses:** Developing tailored security 
protocols for IoT environments, accounting for their unique 
traffic and payload characteristics.

This analysis provides a foundation for developing advanced, adaptive intrusion detection systems (IDS) tailored to IoT environments.



